Data Specialization for a Staged Lambda Calculus

### 2015-05-16 paper/semantics.tex
more stuff in semantics

\subsection{(Small Step) Dynamics}
We provide a small-step semantics which,
given a \lang\ term $e$, evaluates its first stage subterms
either until $e$ is a value (if $e$ is at world \bbonep\ or \bbonem)
of until all of $e$'s first stage subterms have been evaluated away
(when $e$ is at world \bbtwo).
Because first- and second-stage code are
interleaved in \lang\ terms, this first pass of
evaluation requires some care.

Moreover, our staged semantics contains one innovation over those of 
\cite{davies96} and \cite{taha-thesis-99}, which is the addition
of hoisting steps.  
To illustrate, consider the example
\begin{lstlisting}
prev{let x = next{2*3} in next{prev{x}+4}}
\end{lstlisting}
A non-staged semantics which ignores $\next$s and $\prev$s
would proceed on this term by reducing \verb|2+3| to \verb|6| in place.
A stage semantics does not, because it only reduces first stage redexes, and
\verb|2+3| is at the second stage.  
The staged semantics of \cite{davies96} and \cite{taha-thesis-99} instead proceed by
substituting \verb|next{2+3}| in for \verb|x| to yield
\begin{lstlisting}
prev{next{prev{next{2*3}}+4}}
\end{lstlisting}
This takes one step to 
\begin{lstlisting}
prev{next{2*3+4}}
\end{lstlisting}
and then one more step to
\begin{lstlisting}
2*3+4
\end{lstlisting}
which is the final entirely-second-stage result, called a {\em residual}.
Our semantics differs from this formulation in one way.
Consider a slightly modified example where \verb|x| appears twice:
\begin{lstlisting}
prev{let x = next{2*3} in next{prev{x}+prev{x}}}
\end{lstlisting}
A standard stage semantics would, in four steps, evaluate this to 
\begin{lstlisting}
2*3+2*3
\end{lstlisting}
Note how the multiplication is duplicated in the residual.
We find this behavior potentially undesirable for our applications,
so our semantics instead {\em hoists} the contents of a next statement 
to the outermost containing $\prev$, turning
\begin{lstlisting}
prev{let x = next{2*3} in next{prev{x}+prev{x}}}
\end{lstlisting}
into
\begin{lstlisting}
let y = 2*3 in
prev{let x = next{y} in next{prev{x}+prev{x}}}
\end{lstlisting}
This then evaluates to 
\begin{lstlisting}
let y = 2*3 in y+y
\end{lstlisting}
in four steps.  
This hoisting step occurs as the last step in the evaluation of a $\next$.

>>>>

\subsection{(Small Step) Dynamics}
We provide a small-step semantics which,
given a \lang\ term $e$, evaluates its first stage subterms
either until $e$ is a value (if $e$ is at world \bbonep\ or \bbonem)
or until all of $e$'s first stage subterms have been evaluated away
(when $e$ is at world \bbtwo).
Because first- and second-stage code are
interleaved in \lang\ terms, this first pass of
evaluation requires some care.

Moreover, our staged semantics contains one innovation over those of 
\cite{davies96} and \cite{taha-thesis-99}, which is the addition
of {\em hoisting steps}.  
To illustrate, consider the example
\begin{lstlisting}
prev{let x = next{2*3} in next{prev{x}+4}}
\end{lstlisting}
A non-staged semantics which ignores $\next$s and $\prev$s
would proceed on this term by reducing \verb|2*3| to \verb|6| in place.
A staged semantics does not, because it only reduces first stage redexes, and
\verb|2*3| is at the second stage.  
The staged semantics of \cite{davies96} and \cite{taha-thesis-99} instead proceed by
substituting \verb|next{2*3}| in for \verb|x| to yield
\begin{lstlisting}
prev{next{prev{next{2*3}}+4}}
\end{lstlisting}
This takes one step to 
\begin{lstlisting}
prev{next{2*3+4}}
\end{lstlisting}
and then one more step to
\begin{lstlisting}
2*3+4
\end{lstlisting}
which is the final entirely-second-stage result, called a {\em residual}.
Our semantics differs from this formulation in one way.
Consider a slightly modified example where \verb|x| appears twice:
\begin{lstlisting}
prev{let x = next{2*3} in next{prev{x}+prev{x}}}
\end{lstlisting}
A standard stage semantics would, in four steps, evaluate this to 
\begin{lstlisting}
2*3+2*3
\end{lstlisting}
Note how the multiplication is duplicated in the residual.
We find this behavior potentially undesirable for our applications,
so our semantics instead {\em hoists} the contents of a next statement 
to the outermost containing $\prev$, turning
\begin{lstlisting}
prev{let x = next{2*3} in next{prev{x}+prev{x}}}
\end{lstlisting}
into
\begin{lstlisting}
let y = 2*3 in
prev{let x = next{y} in next{prev{x}+prev{x}}}
\end{lstlisting}
This then evaluates to 
\begin{lstlisting}
let y = 2*3 in y+y
\end{lstlisting}
in four steps.  
This hoisting step occurs as the last step in the evaluation of a $\next$.
